---
title: "STAT 331/531: Final Project Analysis"
author: "Jenna Chan, Swara Kurakula, Chirs Li, Wesley Tam"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-fold: true
    toc: true
    number-sections: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
expenditure <- read_csv("data/expenditure_per_student_primary_percent_of_gdp_per_person.csv")
completion <- read_csv("data/primary_completion_rate_total_percent_of_relevant_age_group.csv")
```

## Introduction

Detailed data and variable description. Hypothesized relationship between the variables (and any outside references). Discussion of data cleaning process and decisions.

## Data Cleaning

We chose to limit our analysis to the period from 2006 to 2016, as this range contained the majority of the data with minimal missing values. For each year to be considered acceptable, both expenditure and completion rate needed to be recorded. Any year missing either of these metrics was excluded. Additionally, countries with fewer than 8 acceptable measurements within this timeframe were omitted to ensure that our analysis had a sufficient amount of data for each country.

Our data cleaning process involved transforming the datasets for expenditure per person and primary school completion rates into a long format for the selected years. We then merged these datasets by country and year. And then filtered the data as mentioned earlier.

We also summarized the dataset by calculating the average expenditure and completion rate for each country over the specified period.

```{r}
# Pivot the data
expenditure_longer <- expenditure |>
  select(country, `2006`:`2016`) |>
  pivot_longer(cols = `2006`:`2016`,
               names_to = "year",
               values_to = "expenditure")
completion_longer <- completion |>
  select(country, `2006`:`2016`) |>
  pivot_longer(cols = `2006`:`2016`,
               names_to = "year",
               values_to = "completion_rate")

# Join the datasets
joined <- expenditure_longer |>
  full_join(completion_longer,
            join_by(country == country, year == year))

# Remove countries with too little data
ACCEPTABLE_THRESHOLD <- 8
joined <- joined |>
  mutate(is_acceptable = !is.na(completion_rate) & !is.na(expenditure)) |>
  group_by(country) |>
  mutate(num_acceptable = sum(is_acceptable)) |>
  filter(num_acceptable >= ACCEPTABLE_THRESHOLD & 
         is_acceptable == TRUE)

# Summarize the Data
joined <- joined |>
  summarize(avg_expenditure = mean(expenditure),
            avg_completion = mean(completion_rate))
```
